Deploying and Sharing QSAR Models using Predictive Modeling Markup Language

### 2014-11-25 paper.tex
Three levels of model representation

>>>>

\subsection{Representation of models}

It is possible to distinguish between three levels:
\begin{itemize}
  \item Data structure. Tied to a specific library and/or algorithm.
  For example, the \textit{randomForest} R library returns the trained
  RF model as a special-purpose data structure. Moreover, the internal
  structure of this data structure varies depending on whether the RF
  model was trained using the "formula interface" or the "matrix interface".
  Such data structures can be easily persisted into files using so-called
  serialization mechanisms. The storing operation (ie. saving the in-memory
  data structure to disk file) is called serialization, and the loading
  operation (ie. reading the disk file data structure to memory) is called
  deserialization. Most language runtimes provide built-in serialization
  capabilities. For example, R language provides methods \textit{saveRDS}
  and \textit{readRDS} that can work with binary and text-based files.
  \item Pseudocode. A similar model can be trained using different libraries
  and/or algorithms. For example, a RF model can also be trained using
  the \textit{party} R library (or in-house R library). The choice of the
  algorithm determines how the search is conducted in the feature space. 
  However, the final ending state of a RF training process is an ensemble
  of decision tree models.
  The goal of the pseudocode representation is to extract and encapsulate
  the primary structure of the model. The pseudocode representation aims 
  to be much more informative for humans (eg. following the computation
  path from beginning to end), but still be directly interpretable by 
  machines. For example, a decision tree model could be represented by
  a tree-like data structure, where non-leaf nodes specify split criteria 
  and leaf nodes specify predicted scores.
  \item Application code. During deployment, the model is executed with
  with new input data. With data structure and pseudocode approaches this 
  is accomplished by model interpretation. The alternative approach is to
  translate either representation to application code, which would allow
  for more optimizations and much speedier execution. 
\end{itemize}

Binary serialization mechanisms are typically much more performant and 
compact than text serialization. The only downside, which is rather minor,
is that binary files are hard to open and handle without special tools.
With text files it is possible to take a peek and possibly even make minor
modifications.

The choice of serialization data format depends whether model producer 
and model consumer will be using the same language platform or not. For
example, the RDS data format that is default for R language is not very
suitable when model consumer is a Java or Python application. Today,
there are plenty multi-platform serialization data formats that offer
bindings for most popular programming languages:
\begin{itemize}
  \item Protocol Buffers \url{https://code.google.com/p/protobuf/}
  \item Apache Avro \url{http://avro.apache.org/}
  \item XML
  \item JSON
  \item YAML
\end{itemize}

The first step towards pseudocode would be to have shared data structures.
For example, from the model deployment perspective it would be desirable
if the \textit{randomForest} and \textit{party} libraries were using common
data structure for the representation of decision trees. In practice,
however, it is pretty difficult to accomplish it without a third-party 
standardizing body. The responsibilities of such third party would be 
then to specify, develop and maintain an independent library that could
be imported on demand. The key issue is obtaining acceptance and support
by major tool vendors in the field. Moreover, this library of "common 
data structures" does not have to be limited to one specific language, but
be applicable across all platforms and languages that are used for
everyday statistics and machine learning work.

The predictive analytics community has decided to standardize around
the Predictive Model Markup Language (PMML) [ref?], which is an XML-based
markup language for the representation of predictive solutions.
